Whole-chromosome pairwise alignment using a hierarchical wavefront algorithm

### 2021-05-31 manuscript/main.tex
added the introduction

\newcommand{\Rplus}{\protect\hspace{-.1em}\protect\raisebox{.35ex}{\smaller{\smaller\textbf{+}}}}
\newcommand{\pp}{\Rplus\Rplus}
\newcommand{\Cpp}{\mbox{C\Rplus\Rplus}\xspace}

@@

    \abstract{
        \textbf{Motivation:} Pairwise alignment of sequences is an important step in many bioinformatics analyses,
        including pangenomes building. Pangenomes are sequence models able to provide a full representation of the
        mutual alignment of collections of genomes. The time and memory required to compute a pairwise alignment
        increase quadratically with the sequence length, making it impractical to directly align very long sequences
        without applying heuristic approaches to determine possible syntenic regions in the sequences. Nevertheless,
        with the advances in sequencing technologies, new genome assemblies are produced at a high rate, pressing for
        the development of tools able to align sequences of the order of tens of megabases long.
        \\
        \textbf{Results:} Here we present wfmash, a new gap-affine pairwise aligner designed to align DNA sequences
        at a whole-chromosome scale. wfmash applies a hierarchical implementation of the wavefront alignment
        algorithm to guide the alignment of very long sequences. It scales efficiently to large eukaryotic
        chromosomes, allowing users to perform pairwise alignment of thousands of large genomes using little
        time and memory.
        \\
        \textbf{Availability:} wfmash is written in \Cpp\ programming language. Source code and user manual are freely available at
        \href{https://github.com/ekg/wfmash}{https://github.com/ekg/wfmash}.
        \\
        \textbf{Contact:}
        \href{andreaguarracino@outlook.com}{andreaguarracino@outlook.com},
        \href{erik.garrison@gmail.com}{erik.garrison@gmail.com}
        \\
        \textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.
    }

@@

    Text Text Text Text Text Text Text Text Text Text Text Text Text
    Text Text Text Text Text Text Text Text. Figure~\ref{fig:01}
    shows that the above method Text Text Text Text Text Text Text
    Text Text Text Text Text.  \citep{Bag01} wants to know about
        {\ldots}{\ldots} text follows.
    \begin{equation}
        \sum \text{\it x}+ \text{\it y} =\text{\it Z}\label{eq:01}\vspace*{-10pt}
    \end{equation}

>>>>

\newcommand{\wfmash}{wfmash}
\newcommand{\Rplus}{\protect\hspace{-.1em}\protect\raisebox{.35ex}{\smaller{\smaller\textbf{+}}}}
\newcommand{\pp}{\Rplus\Rplus}
\newcommand{\Cpp}{\mbox{C\Rplus\Rplus}\xspace}

@@

    \abstract{
        \textbf{Motivation:} Pairwise alignment of sequences is an important step in many bioinformatics analyses,
        including pangenomes building. Pangenomes are sequence models able to provide a full representation of the
        mutual alignment of collections of genomes. The time and memory required to compute a pairwise alignment
        increase quadratically with the sequence length, making it impractical to directly align very long sequences
        without applying heuristic approaches to determine possible syntenic regions in the sequences. Nevertheless,
        with the advances in sequencing technologies, new genome assemblies are produced at a high rate, pressing for
        the development of tools able to align sequences of the order of tens of megabases long.
        \\
        \textbf{Results:} Here we present \wfmash\, a new gap-affine pairwise aligner designed to align DNA sequences
        at a whole-chromosome scale. \wfmash\ applies a hierarchical implementation of the wavefront alignment
        algorithm to guide the alignment of very long sequences. It scales efficiently to large eukaryotic
        chromosomes, allowing users to perform pairwise alignment of thousands of large genomes using little
        time and memory.
        \\
        \textbf{Availability:} \wfmash\ is written in \Cpp\ programming language. Source code and user manual are freely available at
        \href{https://github.com/ekg/wfmash}{https://github.com/ekg/wfmash}.
        \\
        \textbf{Contact:}
        \href{andreaguarracino@outlook.com}{andreaguarracino@outlook.com},
        \href{erik.garrison@gmail.com}{erik.garrison@gmail.com}
        \\
        \textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.
    }

@@

    Many biological analyses take advantage of aligning DNA sequences, ranging from read~mapping
    \citep{22388286, BWA_MEM, 23103880} to variant detection~\citep{21478889}, as well as de~novo
    assembly \citep{19251739} and pangenome building~\citep{33177663, 33066802}. Pairwise sequence alignment
    can be applied to identify similar regions that may indicate functional, structural, and/or evolutionary
    relationships between two biological sequences. In this context, pangenomes model the full set
    of genomic elements in a given species or clade~\citep{32453966}. These data structures encode the mutual
    relationships between all the genomes represented, in contrast to reference-based approaches which
    relate sequences to a particular genome, chosen as reference. The unbiased approach allows studying
    the entire genetic diversity of a population. This opportunity, in combination with the massive amount
    of data available, made possible thanks to the advancement in sequencing technology, presses for the
    development of tools able to compute pairwise alignment of very long sequences.

    To face the upcoming challenges in pangenomes building, we have developed \wfmash\, a new gap-affine
    pairwise tool for aligning DNA sequences at a whole-chromosome scale. It supports split-read alignment
    and gap-affine penalties for insertion and deletions. \wfmash\ scales efficiently to large eukaryotic
    chromosomes while requiring little time and memory, making it suitable to be applied on the pairwise
    alignment of hundreds of genomes. \wfmash\ is a DNA sequence aligner based on the integration of
    MashMap~\citep{30423094}, a fast approximate aligner for computing local alignment boundaries between
    long DNA sequences, and the wavefront alignment algorithm (WFA)~\citep{32915952}, an exact gap-affine algorithm
    that takes advantage of homologous regions between the sequences to accelerate the alignment process.

    %TODO Some sentences of why it exists with respect to Minimap2/Winnomap2, explaining the limitations of those tools in the pairwise alignment task.
